Mechanistic Bayesian Forecasts of COVID19

### 2020-08-18 paper.tex
rework intro

As of August 1st 2020 the Covid-19 pandemic has caused over 4.5 million infections and over 150,000 deaths in the United States alone. Effective modeling of the course of the pandemic can help assist with public health resource planning, intervention efforts, and vaccine clinical trials. However, data quality during a pandemic suffers from under-reporting, backlog reporting, and limited testing. Classical infectious disease differential equation models, such as the susceptible-exposed-infected-recovered (SEIR) model, do not take these data issues into account. We propose a novel Bayesian compartmental model (MechBayes) that allows for non-parametric modeling of non-pharamacuetical interventions, time-varying testing, and joint observations on case counts and deaths. The model has been used to submit forecasts to the COVID19-ForecsatHub repository organized by the Reich Lab. We examine the performance relative to a baseline model in addition to performing a nested model comparison of our extensions to the classic SEIR models. We demonstrate a significant gain in both point and probabilistic forecast scoring measures when taking into account data quality issues during a real-time pandemic.
%\begin{itemize}
%\item Covid has affected XX people
%\item Forecasts useful for public health resource planning, intervention planning (vaccine trials), and disease burden
%\item Basic Mechanistic Models unable to capture complexities of real world disease epidemics due to 
%\begin{itemize}
%\item Complexities of interventions 
%\item Issues with testing and reporting
%\end{itemize}
%\item We propose a novel forecasting algorithm to overcome 
%\begin{itemize}
%\item Under-reporting of cases
%\item Time-varying interventions
%\end{itemize}
%\item Bayesian end to end estimation using both cases and deaths in numpyro
%\end{itemize}

@@

The emergence of COVID-19 in early 2020 in the United States developed into the largest pandemic the country has seen in over a century. As of August 1st 2020, there are over 4.5 million confirmed infections and over 150,000 deaths due to COVID-19 \cite{dong2020interactive}.  Understanding the future trajectory of the pandemic is crucial for minimizing the impact across the nation in terms of healthcare burden, economic impact, and clinical response. Forecasts of incident and cumulative deaths due to COVID may help in resource allocation, vaccine clinical trial planning, and re-opening strategies. Along with non-pharamacutical interventions, forecasts are one of the few public health tools available to help fight the pandemic in the absence of a vaccine. Infectious disease forecasting has been demonstrated to benefit public health decision makers during annual influenza outbreaks \cite{lutz2019applying}\cite{myers2000forecasting}.  However, many forecasts of seasonal disease, such as influenza, often rely on ample historical data to look for patterns that can be projected forward into the future. In an emerging pandemic situation, models must be able to fit to limited data. The COVID-19 pandemic has seen a resurgence in the use of differential equation models to explain the underlying transmission of a disease through a population. First introduced by  Kermack and McKendrick, the model assumes the each individual is in one of a mutually exclusive set of compartments, typically either the susceptible, exposed, infected, or recovered compartment \cite{kermack1927contribution}. The model is specified by setting the rates of flow of individuals between compartments. While these models have been used since their inception in the early 20th century, the COVID pandemic represents a unique opportunity to explore their properties in real-time.

Compartmental models have been used to effectively model and forecast disease in non-pandemic situations. These include complex compartmental models for Influenza forecasting \cite{shaman2012forecasting}\cite{osthus2017forecasting}\cite{ong2010real}, even including a retrospective model evaluation of the 1918 Influenza pandemic \cite{hall2007real}. Compartmental models have been used not just in respiratory disease but in Ebola \cite{lekone2006statistical}, Measles \cite{bokler1993chaos}, Dengue \cite{syafruddin2012seir} and a wide variety of other disease.

@@

Emerging pandemics create a unique set of challenges for accurately predicting future deaths using differential equation models. These include, but are not limited to, severe under reporting of cases due to asymptomatic transmission, time-varying testing rates, backlog data dumps, and both the addition and removal of control measures such as social distancing, lockdown, and mask use. In this work, we introduce a set of extensions to the classic compartmental model that are able to account for the real-world and real-time complexities of infectious disease forecasting during a pandemic. We demonstrate the success of the model in both real-time forecast submissions submitted to the Covid19-ForecastHub as well as a nested model comparison to demonstrate the additional forecast accuracy of our extensions beyond the basic SEIR model. In what follows we first describe the available data and forecast submission infrastructure, outline the basic susceptible-exposed-infected-recovered (SEIR) compartment model, describe our extensions for real-world pandemic forecasting, and finally evaluate the model using both real-time evaluation from submissions and a retrospective model component analysis.

>>>>

The COVID-19 pandemic emerged in late December 2019. In the first six months of the global outbreak, the US reported more cases and deaths than any other country in the world.. Effective modeling of the course of the pandemic can help assist with public health resource planning, intervention efforts, and vaccine clinical trials. However, data quality during a pandemic suffers from under-reporting, delayed reporting of cases and deaths due to reporting infrastructure issues, and limited testing. We propose a novel Bayesian compartmental model (MechBayes) that builds upon the classic compartmental framework of susceptible-exposed-infected-recovered (SEIR) model to operationalize Covid-19 in forecasting in real time. This includes non-parametric modeling of non-pharamacuetical interventions, time-varying rates of testing for the disease, and joint observations on new case counts and new deaths. The model has been used to submit forecasts to the US Centers for Disease Control, through the COVID19-ForecsatHub repository organized by the Reich Lab. We examine the performance relative to a baseline model in addition to performing a nested model comparison of our extensions to the classic SEIR models. We demonstrate a significant gain in both point and probabilistic forecast scoring measures using MechBayes.
%\begin{itemize}
%\item Covid has affected XX people
%\item Forecasts useful for public health resource planning, intervention planning (vaccine trials), and disease burden
%\item Basic Mechanistic Models unable to capture complexities of real world disease epidemics due to 
%\begin{itemize}
%\item Complexities of interventions 
%\item Issues with testing and reporting
%\end{itemize}
%\item We propose a novel forecasting algorithm to overcome 
%\begin{itemize}
%\item Under-reporting of cases
%\item Time-varying interventions
%\end{itemize}
%\item Bayesian end to end estimation using both cases and deaths in numpyro
%\end{itemize}

@@

The emergence of COVID-19 in early 2020 in the United States developed into the largest pandemic the country has seen in over a century. Understanding the future trajectory of the pandemic is crucial for minimizing the impact across the nation in terms of healthcare burden, economic impact, treatments and public health response. Forecasts of incident and cumulative deaths due to COVID may help in resource allocation, vaccine clinical trial planning, and re-opening strategies. Forecasts provide important data to decision-makers and the general public and can improve situational awareness of current trends and how they will continue in coming weeks. Infectious disease forecasting, at the time horizon of up to 4 weeks in the future, has benefited public health decision makers during annual influenza outbreaks \cite{lutz2019applying}\cite{myers2000forecasting}.  However, many forecasts of seasonal disease, such as influenza, often rely on ample historical data to look for patterns that can be projected forward into the future. In an emerging pandemic situation, models must be able to fit to limited data. In modeling the COVID-19 pandemic, many research groups have turned to the use of differential equation models to explain the underlying transmission of a disease through a population. First introduced by  Kermack and McKendrick, the model assumes the each individual is in one of a mutually exclusive set of compartments, typically either the susceptible, exposed, infected, or recovered compartment \cite{kermack1927contribution}. The model is specified by setting the rates of flow of individuals between compartments. While these models have been used since their inception in the early 20th century, the COVID pandemic represents a unique opportunity to explore their properties in real-time at both local and global scales .

Compartmental models have been used to effectively model and forecast disease in non-pandemic situations both retrospectively and in real-time. These include complex compartmental models for real-time influenza forecasting \cite{shaman2012forecasting}\cite{osthus2017forecasting}\cite{ong2010real}, even including a retrospective model evaluation of the 1918 influenza pandemic \cite{hall2007real}. Compartmental models have been used not just in respiratory disease but in Ebola \cite{lekone2006statistical}, measles \cite{bokler1993chaos}, dengue \cite{syafruddin2012seir} and a wide variety of other diseases

In this work, we introduce a novel operational forecast model based on mechanistic foundations and tailored to the particular needs and data availability of COVID-19. These include, but are not limited to, severe under reporting of cases due to low testing rates especially in mild or asymptomatic cases, time-varying testing rates, delayed reporting data dumps, and both the addition and removal of control measures such as social distancing, lockdown, and mask use. The case and death data is reported as new cases, instead of prevalent, and new deaths, instead of cumulative, which we handle using an observation model on top of the underlying SEIR dynamics. Finally, we choose a Bayesian framework that allows for uncertainty in the epidemiological parameters that are unidentifiable from the data, introducing flexibility suited to forecasting. This also allows for implementation in a cutting edge probabilistic programming framework for speed and accuracy.

. Our main goal in this work is to forecast observed incident deaths, and because we have limited historical data on COVID-19, we think compartmental models are among the most parsimonious models available for forecasting. However, our m We are not interested in identifying internal parameters of the model, many of which are poorly determined or not identifiable from the available data. We are willing to accept biologically unrealistic parameter values or priors for latent variables; but not ones that lead to gross pathologies during inference and forecasting.

 MechBayes is able to account for these operational concerns when forecasting in real-time. We demonstrate the success of the model in both forecast submissions submitted to the US Centers for Disease Control via the Covid19-ForecastHub as well as an ablation model comparison to demonstrate the additional forecast accuracy of our extensions beyond the basic SEIR model. In what follows we first describe the available data and forecast submission infrastructure, outline the basic susceptible-exposed-infected-recovered (SEIR) compartment model, describe our extensions for real-world pandemic forecasting, and finally evaluate the model using both real-time evaluation from submissions and a retrospective model component analysis.

\section{Related Work}

@@
